---
output: 
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

<!-- badges: start -->
[![R-CMD-check](https://github.com/IMPALA-Consortium/tsoa/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/IMPALA-Consortium/tsoa/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

# Time Series Outliers and Anomalies (TSOA)
Pekka Tiikkainen, Bayer Pharmaceuticals


## Introduction

TSOA is an R package for identifying outliers and anomalies in clinical trial time series.
Its main focus is on flagging sites with one or more study parameters whose profiles differ
from those of the other sites. However, also anomalies for individual subjects can be identified
with the results the package gives.

This document focuses on the data format requirements of the input and output data. For details on
methodology, please refer to the paper and presentation published at the PHUSE EU Connect 22
meeting. The files are attached to this repository.

## How to use the package

First the clinical data has to be wrangled in a format compatible with the main function's parameters
(please see below for definitions). Please note that data preparation step is intentionally out-of-scope
for the package because each user is likely to have their data in a different format.

Then the main function (process_a_study) is called separately for each study you wish to process.
The function returns the results as a list of data frames. For details on the output, please see below.



## Example

```{r example}

library(tsoa)

data("tsoa_data", package = "tsoa")

tsoa_data

feats <- c(
    "autocorr",
    "average",
    "own_site_simil_score",
    "sd",
    "unique_value_count_relative",
    "lof",
    "range"
  ) %>%
  paste(collapse = ";")

ls_tsoa <- process_a_study(
  data = tsoa_data$data,
  subjects = tsoa_data$subjects,
  parameters = tsoa_data$parameters,
  custom_timeseries = tsoa_data$custom_timeseries,
  custom_reference_groups = tsoa_data$custom_reference_groups,
  default_timeseries_features_to_calculate = feats,
  default_minimum_timepoints_per_series = 3,
  default_minimum_subjects_per_series = 3,
  default_max_share_missing_timepoints_per_series = 0.5,
  default_generate_change_from_baseline = FALSE,
  autogenerate_timeseries = TRUE
)

ls_tsoa
```

## Input parameters

The function process_a_study has a number of parameters that need to be provided. Some of these are data frames
while others are simple variables.

The following gives more details on each parameter. Examples are also provided where needed.

### timeseries_features_to_calculate

Vector of features to calculate for each time series.
The vector must contain at least one of the following feature codes:

| Feature code | Description |
| --- | --- |
| autocorr | Auto-correlation |
| average | Average value |
| own_site_simil_score | Measure of co-clustering of time series from the same study site |
| sd | Standard deviation |
| unique_value_count_relative | Number of unique values divided by number of values available |

### default_minimum_timepoints_per_series

Minimum number of time points for auto-generated time series. This value will be used for
a parameter if they do not have the time_point_count_min column defined in the parameters df.

### default_minimum_subjects_per_series

Minimum number of eligible subjects for auto-generated time series. This value will be used for
a parameter if they do not have the subject_count_min column defined in the parameters df.

### default_max_share_missing_timepoints_per_series

Maximum share of missing time points a subject can have in order to be eligible for a time series. 
This value will be used for a parameter if they do not have the max_share_missing defined in the parameters df.

### default_generate_change_from_baseline

If TRUE, change-from-baseline adjusted version of each time series is generated (in addition to the time series
with actual measurements).

### autogenerate_timeseries

If TRUE, time series are autogenerated in a data-driven matter for all parameters. If set to FALSE, the
custom_timeseries data frame must have at least one entry.

### subjects
Data frame with one row per study subject.

Data frame columns:

| Column name | Data type | Data required | Description |
| --- | --- | --- | --- |
| subject_id | chr | Y | Unique identifier for each subject |
| country | chr | Y | Country where the subject is enrolled. |
| site | chr | Y | Name of the study site. |

### parameters

Data frame with one row for each parameter that should be processed.
Parameters are, for example, laboratory assays and vital sign measurements done
during the course of a study.

Data frame columns:

| Column name | Data type | Data required | Description |
| --- | --- | --- | --- |
| parameter_id | chr | Y | Parameter identifier |
| parameter_category_1 | chr |  | Top level category of the parameter (e.g. LB for a laboratory assay). |
| parameter_category_2 | chr |  | Second level category of the parameter (e.g. General chemistry for a laboratory assay). |
| parameter_category_3 | chr |  | Third level category of the parameter (e.g. Local lab for a laboratory assay). |
| parameter_name | chr | Y | Parameter name |
| time_point_count_min | int |  | Minimum number of time points required for auto-generated time series. If not provided, global default value is used. |
| subject_count_min | int |  | Minimum number of eligible subjects required for auto-generated time series. If not provided, global default value is used. |
| max_share_missing | float |  | Maximum number of missing data points an eligible subject can have for auto-generated time series. If not provided, global default value is used. |
| generate_change_from_baseline | boolean |  | Whether to also generate baseline-adjusted time series for the parameter. If not provided, global default value is used. |

Example for two parameters, one vital sign and one local lab. For any time series generated for
the vital sign, at least 50 eligible subjects are required. For the local lab, time series with one
time point only are acceptable. These parameter-specific properties override the corresponding global
parameters (see above).

| parameter_id | parameter_category_1 | parameter_category_2 | parameter_category_3 | parameter_name | time_point_count_min | subject_count_min | max_share_missing | generate_change_from_baseline |
| --- | --- | --- | --- | --- | --- | --- | --- | --- |
| dummy_param_1 	| VS 	| NA 	| NA 	| DIABP 	|  	| 50 	|  	|  |
| dummy_param_2 	| LB 	| General chemistry 	| Local lab 	| Glucose 	| 1 	|  	|  	|  |

### data

Data frame for the data collected during the study with one row per data point.

Data frame columns:

| Column name | Data type | Data required | Description |
| --- | --- | --- | --- |
| subject_id | chr | Y | Pointer to the subject identifier in the subjects table. |
| parameter_id | chr | Y | Pointer to the parameter identifier in the parameters table. |
| timepoint_1_name | chr | Y | Name of the top level time point (e.g. Visit 1). |
| timepoint_2_name | chr |  | Name of the second level time point (e.g. "30 mins after dosing"). |
| timepoint_rank | int | Y | Integer for ranking time points into chronological order. Ranking is separate for each parameter. |
| result | float | Y | Numerical result of the parameter. |
| baseline | float |  | Possible baseline value for the parameter and the subject. This is used to calculate change-from-baseline time series. |

Example of data contents:

| subject_id	| parameter_id	| timepoint_1_name	| timepoint_2_name	| timepoint_rank	| result	| baseline	|
| --- | --- | --- | --- | --- | --- | --- |
| subj_1	| dummy_param_1	| Visit 1	| pre-dose	| 1	| 72	| 65	|
| subj_1	| dummy_param_1	| Visit 1	| post-dose	| 2	| 81	| 65	|
| subj_1	| dummy_param_1	| Visit 2	| NA	| 3	| 69	| 65	|
| subj_1	| dummy_param_1	| Visit 3	| NA	| 4	| 70	| 65	|
| subj_1	| dummy_param_2	| Visit 3	| NA	| 1	| 85	| 84	|
| subj_2	| dummy_param_2	| Visit 3	| NA	| 1	| 102	| 96	|


### custom_timeseries

The study team might be interested in specific time series - for instance in an oncology trial
one might be interested in data collected in the first two treatment cycles.

When time series are auto-generated in a data-driven manner, it cannot be guaranteed that a particular
time series is generated. To make sure that a specific time series is analyzed, this can be
defined with the _custom_timeseries_ data frame.

Please note: if you do not have any custom time series to process, you still have to provide the main
function a blank data frame!

Data frame columns:

| Column name | Data type | Data required | Description |
| --- | --- | --- | --- |
| timeseries_id | chr | Y | Identifier for the the custom time series. |
| parameter_id | chr | Y | Pointer to the parameter identifier in the parameters table. |
| timepoint_combo | chr | Y | Semicolon-delimited string of timepoint ranks that the time series consists of. These must match values in the timepoint_rank column of the data df. |

In the example below, a custom time series has been defined for dummy_param_1.
Timepoints are defined with their respective ranks in the _data_ df.

| timeseries_id	| parameter_id	| timepoint_combo	|
| --- | --- | --- |
| custom_ts_1	| dummy_param_1	| 1;2;3;4	|

## Output data

Output is a list of four data frames. Their details are explained in the following.

### timeseries

Time series processed. Contains both the auto-generated and/or custom time series.

Data frame columns:

| Column name | Data type | Description |
| --- | --- | --- |
| timeseries_id | chr | Unique identifier for the time series. |
| parameter_id | chr | The parameter the time series is for. |
| baseline | chr | Whether the time series values are original measurements or baseline-adjusted. |
| timepoint_combo | chr | Ranks of the time points that constitute the time series. Semicolon delimited string. |
| timepoint_combo_readable | chr | Names of the time points that constitute the time series. Semicolon-delimited string. |
| timepoint_count | int | Number of timepoints in the time series. |

### timeseries_features

Data frame with features calculated for time series.

Data frame columns:

| Column name | Data type | Description |
| --- | --- | --- |
| timeseries_id | chr | Unique identifier of the time series. Points to the timeseries table. |
| subject_id | chr | Subject ID pointing to the input df subjects. |
| feature | chr | Feature code. |
| feature_value | float | Feature value. |

### PCA_coordinates

Data frame with the top two principal components for a time series. These are useful
for the visualization of the similarity of subjects.

Data frame columns:

| Column name | Data type | Description |
| --- | --- | --- |
| timeseries_id | chr | Unique identifier of the time series. Points to the timeseries table. |
| subject_id | chr | Subject ID pointing to the input df subjects. |
| pc1 | float | The first principal component. |
| pc2 | float | The second principal component. |

### site_scores

Biasness scores for sites. The data frame contains a row for each site/time series/feature combination.

Data frame columns:

| Column name | Data type | Description |
| --- | --- | --- |
| timeseries_id | chr | Unique identifier of the time series. Points to the timeseries table. |
| site | chr | Study site for which the score has been calculated. |
| country | chr | Site's country. |
| feature | chr | Time series feature the score is for. |
| pvalue_kstest_logp | float | Negative logarithm of the raw p-value. |
| kstest_statistic | float | Test statistic of the Kolmogorov-Smirnov test. |
| fdr_corrected_pvalue_logp | float | Site score: negative logarithm of the multiple testing corrected p-value. FDR (False Discovery Rate) is used for correction. |
| reference_group | chr | Reference group for the site. "All" means that the site has been compared to all sites in the study. |
| subject_count | int | Number of site subjects who are eligible for the time series. |








